User Privacy in the Public Bitcoin Blockchain

### 2014-07-04 paper.tex
minor corrections

In this paper we first review the basic elements of the Bitcoin system necessary for the subsequent discussion.
Then we discuss the necessity of avoiding address re-use to prevent unnecessary privacy leakage.
After that, we analyze the Blockchain to determine to which extent address reuse occurs in the Bitcoin community.
Disposable addresses do not offer total protection against privacy leakage.
We detail the remaining risks as well as possible solutions.  
\section{Address Reuse}

@@

Transactions are a fundamental element of Bicoin.
Payments require transactions, and these transactions are shared with the network and securely stored in the Blockchain.
Each transaction consumes some inputs and creates some outputs.
The inputs and outputs are worth bitcoins, and for a regular transaction to be valid the total value of the outputs must not exceed the total amount of the inputs.

The outputs of one transaction can be used as inputs of other transactions.
Critically, each output can be used only once.
In this sense, available outputs are just like available money, which can also only be spent once.
Another name for the available outputs is available coins.
The network does not accept transactions that try to spend coins that have been spent before.

@@

In the example we can observe that another important element of Bitcoin are public/private asymmetric cryptographical keys.
The public key is hashed and coded with some redundancy into base58 addresses.
These addresses are alphanumeric chains that can be used to receive funds.
The outputs of a transaction can be sent to an address, which is simply a convenient representation of a public key.

@@

There is no limit in Bitcoin regarding the number of transactions that can use a given address for the outputs.
The owner of the address' private key will be able to spend each of those outputs once.
Another particularity of bitcoin is that there is no practical limit of the number of addresses that can be generated.
There are $2^{160}$ possible addresses because they are generated using RIPE-MD160.
This makes it possible for users to generate new (disposable) addresses for each incoming payment.

A particular form of outputs which is relevant to the discussion later in this work are \emph{change} outputs.
A user willing to send a Bitcoin payment needs to combine in a transaction a number of inputs of value equal or larger than the desired payment.
If the value of the inputs is larger than the output, it is likely that the sending part does not want to loose the difference, also called change.
In order to keep the change, the payer creates a transaction with inputs exceeding the payment value and two outputs: the actual payment and the change.
The payment is sent to the payee and the change is sent to an address controlled by the payer.

@@

\subsection{The Temptation}

The temptation exists of using Bitcoin addresses just like regular bank accounts numbers.
As addresses are used to receive payments, technically speaking only once is needed.
We can use the same address to receive all our payments, just like we can use a single bank account to receive all our payments.
Obviously, in this case, we will also use the same address for all the payments that we make, which is in agreement with the bank account parallelism.

The difference is that our bank account transactions and balance are kept private by our bank.
In contrast, all the transactions and balances in Bitcoin are publicly available to anyone with Internet access.
We could retain our privacy if we were able to keep our Bitcoin addresses private.
Each time that we participate in a transaction, the other party may have the opportunity to link our address to our identity.

Address reuse is discouraged in the original Bitcoin paper \cite{nakamoto2008bpp}.

@@

Even though address reuse might have worrying consequences, the fact is that we can find address reuse in the Blockchain.
We use Obelisk and Libbitcoin to download and query the Blockchain for evidence of address reuse.
Both Obelisk and Libbitcoin are open source projects under heavy development. 
Our source code to generate the data and the plots presented in the paper is also available in github\footnote{https://github.com/jbarcelo/txfillstat}.
We sweep all the transactions in the Bitcoin history and for every address we count how many times it appears as an output of a transaction.

@@

Wikileaks funding campaign is an example of address re-use.
At the time of this writing, Wikileak's donation webpage offers by default a re-used address.
It also offers the donors the possibility of generating a new (disposable) address by simply clicking a button.
The public address makes it possible for everyone to inspect all the details of the transactions involving that address.
A Blockchain explorer website (such as \texttt{blockchain.info}) can be used to browse all those details.
At the time of this writing, Wikileak's public address has received over 3,854 bitcoins in 2216 transactions.
The source addresses for each transaction are also public.

Wikileaks address reuse can unintendedly leak donors' privacy unless they are very cautious.
If the output of a transaction is later sent to Wikileaks in a second transaction, the payer of the first transaction can learn that the payee of the first transaction is a Wikileaks donor by simply inspecting the Blockchain.

@@

It is important to highlight at this point that the Bitcoin protocol does not require that all the inputs belong to the same wallet.
This is simply a common practice in widespread software wallets.
Therefore is not a protocol vulnerability, but an implementation vulnerability.

A second possible attack involves change addresses.
There are techniques to try to infer which of the outputs of a particular transaction is a change output.
For example, in a transaction which presents two outputs and one of them is smaller than all of the inputs, it can be assumed that the small output is a change output.
Again, this is simply an assumption that relies on the way that popular wallets operate today and by no means is an inherent restriction of the protocol.
If a change output of a transaction is identified and later used as an input in another transaction, an attacker may infer that all the inputs of the first and second transaction belong to the same wallet.

@@

The combination of inputs of different users has been termed \emph{CoinJoin} by the Bitcoin community.
In the simplest case, two users combine two inputs to make two payments in a single transaction.
A simple CoinJoin transactions has two or more inputs from two different users, two outputs which are the two payments, and two change addresses for the two payers.
If the inputs are of similar size, a payee cannot determine which inputs belong to the payer.
Similarly, a Blockchain observer cannot link different inputs as belonging to the same user.

@@

We scan the Blockchain for transactions with two or more inputs and four or more outputs as they are possible CoinJoin transactions.
In Fig.~\ref{fig:coinjoin-candidates-matrix} we present results for a number of inputs and outputs lower than 9.
The total number of transactions in the Blockchain that could be CoinJoin transactions is XXXX.

@@

Both \cite{reid2013aab} and \cite{meiklejohn2013fbc} pay attention to bitcoin thefts and in some cases can track the money until a fraction of the theft reaches an exchange.

@@

The public availability of all Bitcoin transactions poses privacy-related challenges on the protocol.
In this work we discuss address re-use and CoinJoin transactions as two key privacy elements in Bitcoin.
The Bitcoin community is advancing in the understanding of privacy threats and developing mitigating techniques.

>>>>

In this paper we first review the basic elements of the Bitcoin system necessary for the subsequent discussion.
Then we discuss the necessity of avoiding address re-use to prevent unnecessary privacy leakage.
After that, we analyze the Blockchain to determine to which extent address reuse occurs in the Bitcoin community.
As disposable addresses do not offer total protection against privacy leakage, we detail the remaining risks as well as possible solutions.  
\section{Address Reuse}

@@

Transactions are a fundamental element of Bicoin.
Payments require transactions, and these transactions are shared with the network and securely stored in the Blockchain.
Each transaction consumes some inputs and creates some outputs.
The inputs and outputs have a value in bitcoins.
For a regular transaction to be valid, it is required that the total value of the outputs does not exceed the total amount of the inputs.

The outputs of one transaction can be used as inputs of later transactions.
Critically, each output can be used only once.
In the Bitcoin jargon, available outputs are also referred to as available coins.
The network does not accept transactions that try to spend coins that have been spent before.

@@

Other important element of Bitcoin are public/private asymmetric cryptographical keys.
The public key is hashed and coded with some redundancy into base58 addresses.
These addresses are alphanumeric chains that can be used to receive funds.
The outputs of a transaction can be sent to an address, which is simply a convenient representation of a public key.

@@

A Bitcoin address can be used used multiple times to receive transaction outputs.
The owner of the address' private key will be able to spend each of those outputs once.
Another particularity of bitcoin is that there is no practical limit on the number of addresses that can be generated.
The address space is 160 bits which means that the chances of address collision can be neglected for practical purposes.
This makes it possible for users to generate new (disposable) addresses for each incoming payment.

There is particular form of output, the \emph{change} output, that is relevant for the later discussion.
A user willing to send a Bitcoin payment needs to combine in a transaction a number of inputs of value equal or larger than the desired payment.
If the value of the inputs is larger than the output, it is likely that the sending part does not want to loose the difference.
The difference between the total value of the inputs and the desired value of the outputs is called the change.
In order to keep the change, the payer creates a transaction with inputs exceeding the payment value and two outputs: the actual payment and the change.
The payment is sent to the payee and the change is sent to an address controlled by the payer.

@@

\subsection{Address Reuse}

It may be tempting to use  Bitcoin addresses like regular bank accounts numbers.
That is, using a single address to receive many, or even all, payments.

The difference is that bank account transactions and balances are kept private by the bank.
In contrast, all the transactions and balances in Bitcoin are publicly available to anyone with Internet access.
In principle, Bitcoin address are pseudonymous and are not necessarily linked to real identities.
In practice, the pseudonymity can be a very thin line of protection and there are many opportunities to link Bitcoin addresses to owners.

When a particular address is linked to its owner, all the transactions related to that address are also linked to the owner.
Address re-use increases the chances that an address is linked to its owner and also increases
The original Bitcoin paper \cite{nakamoto2008bpp} discourages address reuse.

@@

Despite the fact that address re-use was discouraged since the inception of Bitcoin, we can find evidence of address reuse in the Blockchain.
We use Obelisk and Libbitcoin to download and query the Blockchain for evidence of address reuse.
Both Obelisk and Libbitcoin are open source projects under heavy development. 
Our source code to generate the data and the plots presented in the paper is also available in github\footnote{https://github.com/jbarcelo/txfillstat}.
We sweep all the transactions in the Bitcoin history until January 2014 and, for each address, we count how many times it appears as an output of a transaction.

@@

Wikileaks funding campaign is an example of address reuse.
At the time of this writing, Wikileaks donation webpage offers by default a reused address.
It also offers the donors the possibility of generating a new (disposable) address by simply clicking a button.
If the donor uses the default reused address, it is likely that some privacy information is leaked.

The Blockchain contains all the details of the transactions involving the reused public address.
A Blockchain explorer website (such as \texttt{blockchain.info}) can be used to browse all those details.
At the time of this writing, Wikileak's public address has received over 3,858 bitcoins in 2477 transactions.
The source addresses of each incoming transaction are also public.

These source addresses are pseudonymous and are not, in principle, linked to their owners.
Nevertheless, there are several situations in which the association between addresses and owners is possible.
The most obvious example is when the owner publicizes the address in a personal webpage, blog, forum or any other Internet site.
A second example is when the owner uses that address to receive payments.
If the payer knows both the payee and the address, the address is no longer pseudonymous and can be associated to a real identity.
Similarly, if the address owner uses it for a non-anonymous payment, the identity of the owner also becomes linked to the address.

Combining the facts that some of the addresses donating to Wikileaks are public and some of the owners of those addresses are known, the result is that the identity of some Wikileaks donors is leaked. 
This is not a problem if the donors are aware of this situation and know that Bitcoin, on its own, does not protect the privacy of its users.

@@

It is important to highlight at this point that the Bitcoin protocol does not require that all the inputs belong to the same wallet.
This is simply a common practice in widespread software wallets.
Therefore it is not a protocol vulnerability, but an implementation vulnerability.

A second possible attack involves change addresses.
There are techniques to try to infer which of the outputs of a particular transaction is a change output.
For example, in a transaction which presents two outputs and one of them is smaller than any of the inputs, it can be assumed that the small output is a change output.
Again, this is simply an assumption that relies on the way that popular wallets operate today and by no means is an inherent restriction of the protocol.
If a change output of a transaction is identified and later used as an input in another transaction, an attacker may infer that all the inputs of the first and second transaction belong to the same wallet.

@@

The combination of inputs of different users has been termed \emph{CoinJoin} by the Bitcoin community.
In the simplest case, two users combine two inputs to make two payments in a single transaction.
A simple CoinJoin transactions has two or more inputs from two different users, two outputs which are the two payments, and two change addresses for the two payers.
If the inputs are of similar size, a payee cannot determine which inputs belong to which payer.
Similarly, a Blockchain observer cannot link different inputs as belonging to the same user.

@@

We scan the Blockchain for transactions with two or more inputs and four or more outputs as they are possible CoinJoin transactions.
In Fig.~\ref{fig:coinjoin-candidates-matrix} we present results for a number of inputs and outputs lower than 9.

@@

Both \cite{reid2013aab} and \cite{meiklejohn2013fbc} pay attention to bitcoin thefts and, in some cases, can track the money until a fraction of the theft reaches an exchange.

@@

The public availability of all Bitcoin transactions poses privacy-related challenges on the protocol.
In this work we discuss address re-use and CoinJoin transactions as two key privacy elements in Bitcoin.
We present data regarding address re-use in the blockchain as well as possible CoinJoin transactions.
The Bitcoin community is advancing in the understanding of privacy threats and developing mitigating techniques.
